Precise spiking motifs in neurobiological and neuromorphic data

### 2022-12-19 manuscript.tex
Update on Overleaf.

Although surprising, this speed of the visual cortex in primates is compatible with the latencies that are recorded at the neuro-physiological level. The rapid propagation of the visual information in the thalamus, then in the primary visual cortex takes about $45~\ms$ in the macaque~\citep{schmolesky_signal_1998} and about $60~\ms$ in humans~\citep{vanni_coinciding_2001}. This functioning of visual processing as a forward pass is most prominent in fast processing (see Figure~\ref{fig:thorpe}), and can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}. An important consequence of this speed of processing of vision is that it implies that processing is carried out using only very few spikes per layer. As a comparison, the latencies in macaque monkeys are about: Retina, $20-40~\ms$; V1, $40-60~\ms$; IT, $80-100~\ms$; MC, $140-190~\ms$; to finger muscles, $180-260~\ms$. Indeed, we may consider that a behavioral response occurring in only $200~\ms$ would involve about ten processing stages along the ``forward'' pathways of the visual system. Such processes were indeed efficiently reproduced in feed-forward models trained with back-propagation~\citep{serre_feedforward_2007,jeremie_ultrafast_2022}. At the same time, it has been demonstrated that one spike requires a significant amount of time (about $10~\ms$) to be conducted from one layer to the next~\citep{nowak_timing_1997,thorpe_seeking_2001}.  This figure is inspired by similar schematics done for monkeys in~\citep{thorpe_seeking_2001}. As a consequence, these results suggest that ``like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How precise spike timing may encode vectors of real values}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% reliability of spike timings
Sensory data is most often represented by continuous values, such as the energy produced by a flow of photons that hits the different photoreceptors of the retina. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate of cortical cells in awake monkeys is highly irregular~\citep{softky_highly_1993}, which makes it at first sight inconsistent with the temporal integration of firing rate. Remarkably, it has been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible: while the first spike is aligned to stimulus' time, the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{bryant_spike_1976,mainen_reliability_1995} (see Figure~\ref{fig:mainen}). However, if that same neuron is now driven by a \emph{frozen} noise, that is, a highly dynamic signal which is repeatedly replayed from trial to trial, then the output spikes are highly reproducible (for a review, see~\citep{ermentrout_reliability_2008}). This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''

>>>>

Although surprising, this speed of the visual cortex in primates is compatible with the latencies that are recorded at the neuro-physiological level. The rapid propagation of the visual information in the thalamus, then in the primary visual cortex takes about $45~\ms$ in the macaque~\citep{schmolesky_signal_1998} and about $60~\ms$ in humans~\citep{vanni_coinciding_2001}. This functioning of visual processing as a forward pass is most prominent in fast processing (see Figure~\ref{fig:thorpe}), and can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}. An important consequence of this speed of processing of vision is that it implies that processing is carried out using only very few spikes per layer. As a comparison, the latencies in macaque monkeys are about: Retina, $20-40~\ms$; V1, $40-60~\ms$; IT, $80-100~\ms$; MC, $140-190~\ms$; to finger muscles, $180-260~\ms$.\add[Laurent]{Note that, since maximal conduction speeds are roughly constant, theses latencies are comparable to that found in humans, with a ratio given by the physical size of the whole system.} It follows that if we consider that a behavioral response occurra in only $200~\ms$, it would involve about ten processing stages along the ``forward'' pathways of the visual system. Such processes were indeed efficiently reproduced in feed-forward models trained with back-propagation~\citep{serre_feedforward_2007,jeremie_ultrafast_2022}. At the same time, it has been demonstrated that one spike requires a significant amount of time (about $10~\ms$) to be conducted from one layer to the next~\citep{nowak_timing_1997,thorpe_seeking_2001}.  This figure is inspired by similar schematics done for monkeys in~\citep{thorpe_seeking_2001}. As a consequence, these results suggest that ``like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How precise spike timing may encode vectors of real values}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\add[Laurent]{Let's now focus on one processing step along a cortical pathway. }
% reliability of spike timings
Sensory data is most often represented by continuous values, such as the energy produced by a flow of photons that hits the different photoreceptors of the retina. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate of cortical cells in awake monkeys is highly irregular~\citep{softky_highly_1993}, which makes it at first sight inconsistent with the temporal integration of firing rate. Remarkably, it has been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible: while the first spike is aligned to stimulus' time, the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{bryant_spike_1976,mainen_reliability_1995} (see Figure~\ref{fig:mainen}). However, if that same neuron is now driven by a \emph{frozen} noise, that is, a highly dynamic signal which is repeatedly replayed from trial to trial, then the output spikes are highly reproducible (for a review, see~\citep{ermentrout_reliability_2008}). This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''
